\textbf{\sc Towards Deliberative Control in Marine Robotics}

### 2012-06-05 final-version/DeliberativeControl.tex
Final Version to be sent to Publisher

\tableofcontents{}

@@

Rajan and Py are funded by a block grant from the David and Lucile
Packard Foundation to MBARI. They thank their collaborators John Ryan,
Julio Harvey, Jim Bellingham and Robert Vrijenhoek at MBARI. They
thank other members of the \eu team over the years, including Conor
McGann, Ari J\'onsson, Jeremy Frank, Paul Morris, John Bresina,
Michael Iatauro, Matthew Boyce and Tristan Smith.  The authors are
grateful to Martha del Alto, David Korsmeyer and Steven Zornetzer all
at NASA Ames \kcomment{who not only made \eu available as open source
  but encouraged its use outside the agency}. Minh Bin Do (NASA Ames),
William Yew Teck Tan (NUS, Singapore) and Rishi Graham (MBARI)
provided valuable comments on early drafts; we are grateful for their
help in making this manuscript more accessible. Finally, the authors
are indebted to Nicola Muscettola \kcomment{(Google)} for his ideas
and early efforts in Constraint-based Temporal Planning while at NASA
Ames, and which forms the basis of much of this work.

>>>>

% \tableofcontents{}

@@

Rajan and Py are funded by a block grant from the David and Lucile
Packard Foundation to MBARI. They thank their research collaborators
John Ryan, Julio Harvey, Jim Bellingham and Robert Vrijenhoek at
MBARI. The crew of the R/V \emph{Zephyr} and the DMO support team led
by Hans Thomas has been invaluable. The authors also thank other
members of the \eu team over the years, including Conor McGann, Ari
J\'onsson, Jeremy Frank, Paul Morris, John Bresina, Michael Iatauro,
Matthew Boyce and Tristan Smith.  They are grateful to Martha del
Alto, David Korsmeyer and Steven Zornetzer all at NASA Ames
\kcomment{who not only made \eu available as open source but
  encouraged its use outside the agency}. Minh Bin Do (NASA Ames),
William Yew Teck Tan (NUS, Singapore) and Rishi Graham (MBARI)
provided valuable comments on early drafts; we are grateful for their
help in making this manuscript more accessible. Finally, the authors
are indebted to Nicola Muscettola \kcomment{(Google)} for his ideas
and early efforts in Constraint-based Temporal Planning while at NASA
Ames which forms the basis of much of this work.

### 2012-06-05 final-version/concepts.tex
Final Version to be sent to Publisher

\paragraph {A brief note on \eu history and notation} The origin of
what we will call \eu in this chapter, is derived from the
\texttt{HSTS} planner \cite{mus94} used on a number of ground-based
demonstrations for space telescope scheduling as noted in Section
\ref{sec:rabeyond}. \texttt{HSTS} was also the \texttt{LISP}-based
planner that was flown as part of the Remote Agent \texttt{RAX}
experiment. Concurrent to \texttt{RAX} deployment, a C++ version of
the planner was implemented \kcomment{at NASA Ames. This} went on to
be used for the \texttt{MAPGEN} \cite{bresina05} system and continues
to be used to date, in the mission-critical uplink process for the
command/control of the Mars Exploration Rovers (\texttt{MER})
mission. A vastly re-factored, higher performance open source version
\cite{europapso} was named \eut and is the subject of this chapter and
used by \rxe.  Note however, in using the name \eu, we refer to the
\emph{fundamental ideas as well as the planning paradigm} even if
there exists a specific implementation.

@@

\begin{figure}[!htb]
\centering
\includegraphics[scale=2.3]{figs/flexible-timelines.jpeg}
\caption{\small Tokens with flexible temporal intervals and parametric
  constraints between tokens. This example shows the triggering of a
  water sampler based on a feature threshold while the vehicle
  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
  start \& end times.}
\label{fig:flex-timelines}
\end{figure}

@@

\item[\textbf{Variables}] Values that need to be represented to
  describe the problem domain \kcomment{over which constraints are
    specified}.  In the Shopping Agent example, the times at which the
  agent needs to leave or be back, or executes a purchase, are all
  instances where \emph{variables} representing time is used.  As
  noted in Section \ref{sec:europa:cp}, a variable can take values
  from a domain. In \eu, domains are defined over a specific data
  type; the primitive data types supported are \textit{String},
  \textit{Boolean}, \textit{Numeric} and \textit{Object} (see
  below). Further, two types of variable Domains can be defined over a
  data type \textit{Enumerated} and \textit{Interval} (used only for
  numeric data types, they are specified by a
  $[LowerBound,UpperBound]$ pair, for instance: $[1,10]$,
  $[5.0,100.0]$). It is this latter ability to represent and reason
  over numeric intervals that forms the basis of representing flexible
  plans.

@@

\item[\textbf{Tokens}] In first order logic, a predicate defines a
  relation between objects and properties \cite{gen87}.  In \eu, we
  define such relations between variables whose domains are sets of
  objects and sets of properties to describe state and behavior. For
  example, one might use a predicate \texttt{At($a$,$l$)} to indicate
  that agent $a$ is at location $l$, or a predicate
  \texttt{Buy($a$,$p$}) to indicate that agent $a$ is acting to buy
  product $p$. Note that $a$ is a variable which may have a number of
  possible values in a problem with multiple agents. Similarly, $l$
  and $p$ are variables whose values are the set of possible positions
  and products respectively. \eu can be used to create partial plans,
  where the domains of variables $a$ and $l$ can have more than one
  possible value. Or they can be grounded \kcomment{representations},
  where single values will be specified for each variable as shown in
  the $N$-Queens problem.

  In general, for an executable plan it is insufficient to
  \kcomment{define} predicates that describe required state or
  behavior without also specifying some temporal extent over which
  each of those predicates hold. Predicates that are always true can
  be thought to hold from the beginning to the end of the planning
  horizon. However, in practice, the temporal extent of interest must
  be defined with timepoints to represent its start and end.  In such
  \kcomment{a} representation, one writes \texttt{At($a$,$s$,$e$,$l$)}
  to indicate that agent $a$ is at location $l$ from time $s$ to time
  $e$. In fact, this pattern of \kcomment{usage} to describe both
  state and behavior of objects is prevalent enough that \eu
  \kcomment{encapsulates into} the token
  representation. \kcomment{Tokens} have built in variables to
  indicate the object to which \kcomment{it primarily} applies as well
  as the timepoints over which it holds. A token, therefore, is an
  instance of a predicate that represents an object's state or
  behavior and is defined over a temporal extent. In \eu, all
  predicate instances are tokens. \kcomment{In} the same way that
  objects are described generically by classes, in \eu tokens can be
  described generically by token types. Every token has five built-in
  variables:

>>>>

\paragraph {A brief note on \eu history and notation} The origin of
what we will call \eu in this chapter, is derived from the
\texttt{HSTS} planner \cite{mus94} used on a number of ground-based
demonstrations for space telescope scheduling as noted in Section
\ref{sec:rabeyond}. \texttt{HSTS} was also the \texttt{LISP}-based
planner that was flown as part of the Remote Agent \texttt{RAX}
experiment. Concurrent to \texttt{RAX} deployment, a C++ version of
the planner was implemented \kcomment{at NASA Ames. This} went on to
be used for the \texttt{MAPGEN} \cite{bresina05} system and continues
to be used to date, in the mission-critical uplink process for the
command/control of the Mars Exploration Rovers (\texttt{MER})
mission. A vastly re-factored, higher performance open source version
\cite{europapso} was named \eut and is the subject of this chapter and
used by \rxe.  Note however, in using the name \eue, we refer to the
\emph{fundamental ideas as well as the planning paradigm} even if
there exists a specific implementation.

@@

\begin{figure}[!tb]
\centering
\includegraphics[scale=2.3]{figs/flexible-timelines.jpeg}
\caption{\small Tokens with flexible temporal intervals and parametric
  constraints between tokens. This example shows the triggering of a
  water sampler based on a feature threshold while the vehicle
  Yo-Yo's. The \texttt{Waypoint\_Yo-Yo} token has a flexible duration,
  start \& end times.}
\label{fig:flex-timelines}
\end{figure}

@@

\item[\textbf{Variables}] Values that need to be represented to
  describe the problem domain \kcomment{over which constraints are
    specified}.  In the Shopping Agent example, the times at which the
  agent needs to leave or be back, or executes a purchase, are all
  instances where \emph{variables} representing time is used.  As
  noted in Section \ref{sec:europa:cp}, a variable can take values
  from a domain. In \eue, domains are defined over a specific data
  type; the primitive data types supported are \textit{String},
  \textit{Boolean}, \textit{Numeric} and \textit{Object} (see
  below). Further, two types of variable Domains can be defined over a
  data type \textit{Enumerated} and \textit{Interval} (used only for
  numeric data types, they are specified by a
  $[LowerBound,UpperBound]$ pair, for instance: $[1,10]$,
  $[5.0,100.0]$). It is this latter ability to represent and reason
  over numeric intervals that forms the basis of representing flexible
  plans.

@@

\item[\textbf{Tokens}] In first order logic, a predicate defines a
  relation between objects and properties \cite{gen87}.  In \eue, we
  define such relations between variables whose domains are sets of
  objects and sets of properties to describe state and behavior. For
  example, one might use a predicate \texttt{At($a$,$l$)} to indicate
  that agent $a$ is at location $l$, or a predicate
  \texttt{Buy($a$,$p$}) to indicate that agent $a$ is acting to buy
  product $p$. Note that $a$ is a variable which may have a number of
  possible values in a problem with multiple agents. Similarly, $l$
  and $p$ are variables whose values are the set of possible positions
  and products respectively. \eu can be used to create partial plans,
  where the domains of variables $a$ and $l$ can have more than one
  possible value. Or they can be grounded \kcomment{representations},
  where single values will be specified for each variable as shown in
  the $N$-Queens problem.

  In general, for an executable plan it is insufficient to
  \kcomment{define} predicates that describe required state or
  behavior without also specifying some temporal extent over which
  each of those predicates hold. Predicates that are always true can
  be thought to hold from the beginning to the end of the planning
  horizon. However, in practice, the temporal extent of interest must
  be defined with timepoints to represent its start and end.  In such
  \kcomment{a} representation, one writes \texttt{At($a$,$s$,$e$,$l$)}
  to indicate that agent $a$ is at location $l$ from time $s$ to time
  $e$. In fact, this pattern of \kcomment{usage} to describe both
  state and behavior of objects is prevalent enough that \eu
  \kcomment{encapsulates into} the token
  representation. \kcomment{Tokens} have built in variables to
  indicate the object to which \kcomment{it primarily} applies as well
  as the timepoints over which it holds. A token, therefore, is an
  instance of a predicate that represents an object's state or
  behavior and is defined over a temporal extent. In \eue, all
  predicate instances are tokens. \kcomment{In} the same way that
  objects are described generically by classes, in \eu tokens can be
  described generically by token types. Every token has five built-in
  variables:

### 2012-06-05 final-version/europa-basics.tex
Final Version to be sent to Publisher

Planning as an activity is ingrained in human decision-making
\cite{berthoz} and its generalization into a computational framework
demonstrating machine intelligence \kcomment{has for long been}
targeted by pioneers in computer science
\cite{computersthought}. Early work in this context focused on logical
theorem proving by computer \cite{green69} which has formed the basis
of much of the discipline. The emphasis in these early efforts was to
logically derive the existence proof of a trajectory of actions for a
hypothetical robot (moving blocks, chess pieces, furniture etc) to
achieve a specific goal given specific initial conditions. Two
important properties of \emph{Soundness}\footnote{A planning algorithm
  is sound if invoked on a problem $P$ returns a plan which is a
  solution for $P$.} and \emph{Completeness}\footnote{A planning
  algorithm is complete if invoked on a solvable problem $P$ is
  guaranteed to return a solution.} along with derived results for
algorithmic complexity \cite{gareyjohnson,corman} were a defining
cornerstone for understanding how different computational paradigms
could actually be tractable first in simulation and subsequently in
actual robotic platforms. In this chapter our emphasis is to focus on
embedded AI Planning and the execution of plans on an embedded
\kcomment{marine robotic} platform. The techniques we highlight are
general purpose as noted in Section \ref{sec:intro}.

@@

These drawbacks have been addressed by a number of frameworks like
\texttt{Descartes} \cite{joslin95,joslin96}, \texttt{IxTet}
\cite{ghallab87,lemai04} and \texttt{CAIP} (Constraint-Based Interval
and Attribute Planning) \cite{mus94,frank2003}.  They take advantage
of a \textsf{CP} representation to deal with time and resources,
\kcomment{while} maintaining an explicit representation of the
elements relevant to planning (action choices, causal relationships,
etc). This combination results in a planning approach that can deal
with the \kcomment{kind} of constraints found in real-world problems
while allowing for intuitive problem descriptions and a natural way to
express search algorithms and heuristics. \eu is one such
implementation of the \texttt{CAIP} framework, \kcomment{others
  include \cite{chien00, iaai-CestaCFO09}}. In \eu planning problems
are specified in terms of:

>>>>

Planning as an activity is ingrained in human decision-making
\cite{berthoz} and its generalization into a computational framework
demonstrating machine intelligence \kcomment{has for long been}
targeted by pioneers in computer science
\cite{computersthought}. Early work in this context focused on logical
theorem proving by computer \cite{green69} which has formed the basis
of much of the discipline. The emphasis in these early efforts was to
logically derive the existence proof of a trajectory of actions for a
hypothetical robot (moving blocks, chess pieces, furniture etc) to
achieve a specific goal given specific initial conditions. Two
important properties of \emph{Soundness}\footnote{A planning algorithm
  is \emph{sound} if invoked on a problem $P$ returns a plan which is
  a solution for $P$.} and \emph{Completeness}\footnote{A planning
  algorithm is \emph{complete} if invoked on a solvable problem $P$ is
  guaranteed to return a solution.} along with derived results for
algorithmic complexity \cite{gareyjohnson,corman} were a defining
cornerstone for understanding how different computational paradigms
could actually be tractable first in simulation and subsequently in
actual robotic platforms. In this chapter our emphasis is to focus on
embedded AI Planning and the execution of plans on an embedded
\kcomment{marine robotic} platform. The techniques we highlight are
general purpose as noted in Section \ref{sec:intro}.

@@

These drawbacks have been addressed by a number of frameworks like
\texttt{Descartes} \cite{joslin95,joslin96}, \texttt{IxTeT}
\cite{ghallab87,lemai04} and \texttt{CAIP} (Constraint-Based Interval
and Attribute Planning) \cite{mus94,frank2003}.  They take advantage
of a \textsf{CP} representation to deal with time and resources,
\kcomment{while} maintaining an explicit representation of the
elements relevant to planning (action choices, causal relationships,
etc). This combination results in a planning approach that can deal
with the \kcomment{kind} of constraints found in real-world problems
while allowing for intuitive problem descriptions and a natural way to
express search algorithms and heuristics. \eu is one such
implementation of the \texttt{CAIP} framework, \kcomment{others
  include \cite{chien00, iaai-CestaCFO09}}. In \eu planning problems
are specified in terms of:
